processing is split across small steps (df.1, df.2, ... df.22 for easier interpretation)  

```{r}
library(DBI)
require(tidyverse)
```

load MIMIC-III derived and main data from physionet, requires physionet account
see https://physionet.org/content/mimiciii/1.4/ and click on "Request access using Google BigQuery" 
derived data contains a lot of useful abstractions like height, weight, blood pressure, elixhauser scores, etc. we don't need to work with raw data or ICD9 codes directly
connection to derived MIMIC data 

```{r}
con.1 <- dbConnect(  
  bigrquery::bigquery(),
  project = "physionet-data",
  dataset = "mimiciii_derived",
)
```

connection to primary MIMIC data

```{r}
con.2 <- dbConnect(
  bigrquery::bigquery(),
  project = "physionet-data",
  dataset = "mimiciii_clinical"
)
```

Auth bigquery using gmail

```{r}
bigrquery::bq_auth(email = "EMAIL") 
```

Read tables with metrics required for analysis 

```{r}
hw.iii=dbReadTable(con.1,'heightweight') #' height and weight data
icu.iii=dbReadTable(con.2,'icustays') ## icu stays 
elix.iii=dbReadTable(con.1,'elixhauser_quan_score') ## elixhauser score
elix.iii.2=dbReadTable(con.1,'elixhauser_quan') ## elixhauser score, including depression, substance abuse
echo.iii=dbReadTable(con.1,'echodata_structured') ## echo data including blood pressure
labevents=dbReadTable(con.2,'labevents')    #' load all lab events
a1c=labevents %>% filter(ITEMID==50852|ITEMID==50854)  ## A1C from lab events
c.protein=labevents %>% filter(ITEMID==50889)  ## c protein from lab events
chol=labevents %>% filter(ITEMID==50907) #' cholestorol from lab events
```

These 3 MIMIC-iii files (patients, admissions, textnotes) can be downloaded from physionet https://physionet.org/content/mimiciii/1.4/

```{r}
pat=read.csv('PATIENTS.csv')
adm=read.csv('ADMISSIONS.csv')
not=read.csv('NOTEEVENTS.csv')
```

we will only use discharge summaries from notes corpus

```{r}
df.1=
  not %>% tidytable::filter(CATEGORY=='Discharge summary')  
```

extract social history section of the note with reg exp, we try different matches like all caps, and others, and then combine them.

```{r}
df.2=
  df.1 %>% 
  tidytable::mutate(text=str_match(tolower(TEXT) %>% gsub('[\r\n]',' ',.),'social history.[ ]*(.*)family history')[,2]) %>% 
  filter(!is.na(text))

df.2.other=
df.1 %>% 
  tidytable::mutate(text=str_match(tolower(TEXT) %>% gsub('[\r\n]',' ',.),'social history.[ ]*(.*)family history')[,2]) %>% 
  filter(is.na(text)) 

df.2.other.2=
  df.2.other %>%
  tidytable::mutate(text=str_match(TEXT %>% gsub('[\r\n]',' ',.),'SOCIAL HISTORY.[ ]*(.*?)[A-Z]+:')[,2]) %>% 
  filter(!is.na(text))

df.2.other.3=
  df.2.other %>%
  tidytable::mutate(text=str_match(TEXT %>% gsub('[\r\n]',' ',.),'Social History.[ ]*(.*?)[A-Z]+:')[,2]) %>% 
  filter(!is.na(text))
df.2.all=bind_rows(df.2,df.2.other.2,df.2.other.3) %>% mutate(TEXT=text)
```

now we will annotate notes with living with label
helper function 

```{r}
get.maxscore=function(text,pipe=pipe)
{
  p=pipe(text)
  print(p)
  (p %>% unlist)[which((p %>% unlist %>% names) =='score')] %>% which.max
}
```

we will have to use reticulate to use UC NLP models, we will need python with prerequisites: transformers, spacy
see documentation for UC CNN and bag of words (BOW) models for dependencies and installation
https://github.com/BCHSI/social-determinants-of-health-clbp

```{r}
reticulate::use_python('/usr/bin/python3')
library(reticulate)
library(tidyverse)
mod='hf.model.lw.shac' ## SHAC model
```

install python transformers before running next line

```{r}
tf=import('transformers')
m=tf$AutoModelForSequenceClassification$from_pretrained(mod)
t=tf$AutoTokenizer$from_pretrained(mod)
pipe = tf$TextClassificationPipeline(model=m, tokenizer=t,padding="max_length",truncation=T, return_all_scores=T)

levels=c("family", "others", "spouse", "alone", "partner", "not given")
```

annotate notes, long process

```{r}
df.3=df.2.all %>% rowwise %>% mutate(label=list(get.maxscore(TEXT,pipe)))
```

get factor representation instead of numbers for lw (living with) label

```{r}
df.4=df.3 %>% ungroup %>% mutate(lw.label=factor(unlist(label),levels=1:length(levels),
                                                             labels=c(levels)))

saveRDS(df.4,'df.4')
```

now we will use spacy via reticulate to apply CNN model for extraction of marital status

```{r}
spacy=import('spacy')
nlp=spacy$load("en_sdoh_cnn_ner_cui")
```

this function applies NLP model to texts in parallel to extract label_ (probably shorter function could be written but I am not expert in reticulate/python combo)

```{r}
nlp_extr.all=function(texts){
  docs=nlp$pipe(texts)
  ap=function(i) {
    l=list()
    for (ent in i$ents)
      l=append(l,list(c(
        ent$label_ )))
    return(l)
  }
  return(iterate(docs,ap,simplify = F) )
}
```

now perform annotation, long process ~several hours

```{r}
df.5=df.4 %>% mutate(extr=nlp_extr.all(TEXT))
```

annotations come in complex format, which we will need to work with to extract the actual label

```{r}
df.6=
  df.5 %>% 
  rowwise %>% 
  mutate(extr.na=max(ifelse(grepl("Marital",extr),1,0))) %>% 
  ungroup %>% 
  unnest(extr,keep_empty = T) %>% 
  unnest_wider(extr,names_sep='_')  %>% 
  separate(extr_1,sep=':',into=c('category','label')) 
```

processing continues, we also want to only use marital status, since model extracts many other label categories too

```{r}
df.7=
  df.6 %>% 
  mutate(label=ifelse(is.infinite(extr.na)|extr.na==0,'not given',label)) %>% 
  mutate(category=ifelse(is.infinite(extr.na)|extr.na==0,'Marital_or_partnership_status',category)) %>% 
  filter(category %in% c("Marital_or_partnership_status")) %>% 
  group_by(HADM_ID,category,TEXT) %>% 
  mutate(label=ifelse(!all(is.na(label)),tail(names(sort(table(label))), 1),label)) %>%  #' in case one text gets several marital status predictions, we take the most frequent occuring label
  slice(1) %>% 
  ungroup %>% 
  pivot_wider(names_from=category,values_from=label)  
```

set names cnn.label will be marital status extracted by CNN

```{r}
df.8=
  df.7 %>% 
  rename(cnn.label="Marital_or_partnership_status") %>% 
  mutate(cnn.label=trimws(tolower(cnn.label))) 
```

Save for faster access

```{r}
saveRDS(df.8,'df.8')
```

We are done with CNN model, now we can extract marital status using Bag of words model (BOW)

```{r}
nlp=spacy$load("en_sdoh_bow_cui") #' restart r session, if reticulate function doesn't work with other model in cache
```

annotate, long process

```{r}
df.9=readRDS('df.8') %>% mutate(extr=nlp_extr.all(TEXT))
```

do similiar processing for BOW model as for CNN

```{r}
df.10=
  df.9 %>% 
  rowwise %>% 
  mutate(extr.na=max(ifelse(grepl("Marital",extr),1,0))) %>% 
  ungroup %>% 
  unnest(extr,keep_empty = T) %>% 
  unnest_wider(extr,names_sep='_')  %>% 
  separate(extr_1,sep=':',into=c('category','label'))

df.11=
  df.10 %>% 
  mutate(label=ifelse(is.infinite(extr.na)|extr.na==0,'not given',label)) %>% 
  mutate(category=ifelse(is.infinite(extr.na)|extr.na==0,'Marital_or_partnership_status',category)) %>% 
  filter(category %in% c("Marital_or_partnership_status")) %>% 
  group_by(HADM_ID,category,TEXT) %>% 
  mutate(label=ifelse(!all(is.na(label)),tail(names(sort(table(label))), 1),label)) %>% 
  slice(1) %>% 
  ungroup %>% 
  pivot_wider(names_from=category,values_from=label)  
```

set names, hybrid.label will be BOW model marital status 

```{r}
df.12=
  df.11 %>% 
  rename(hybrid.label="Marital_or_partnership_status") %>% 
  mutate(hybrid.label=trimws(tolower(hybrid.label))) 
```

now an important step we recode partnered patients using lw label. 
We consider partnered only patients who live with their partner, girlfriend, boyfriend or significant other
All other partnered are considered dating

```{r}
df.13= 
  df.12 %>% 
  select(-extr.na) %>% 
  mutate(across(contains("label"),function(x) case_when(x=='single person'~'single',   #' minor recoding of marital status labels is done with case_when
                                                        x=='life partner'~'partner',
                                                        x=='na'~'not given',
                                                        is.na(x)~'not given',
                                                        x==''~'not given',
                                                        x=='unknown (default)'~'not given',
                                                        .default=x))) %>%
  mutate(hybrid.label=case_when(
    hybrid.label=="partner"&lw.label=="partner"~hybrid.label,
    hybrid.label=="partner"&lw.label=="family"~"dating",
    hybrid.label=="partner"&lw.label=="others"~"dating",
    hybrid.label=="partner"&lw.label=="alone"~"dating",
    .default = hybrid.label
  )) %>% mutate(cnn.label=case_when(
    cnn.label=="partner"&lw.label=="partner"~cnn.label,
    cnn.label=="partner"&lw.label=="family"~"dating",
    cnn.label=="partner"&lw.label=="others"~"dating",
    cnn.label=="partner"&lw.label=="alone"~"dating",
    .default = cnn.label 
  )) 
```

Now we do majority voting, if 2 out 3 votes match on marital status we use the label, otherwise we set to NA

```{r}
df.14=
  df.13 %>% 
  left_join(adm,by=c("SUBJECT_ID","HADM_ID")) %>% 
  mutate(label=case_when(
    cnn.label==hybrid.label~hybrid.label,
    cnn.label==MARITAL_STATUS~MARITAL_STATUS,
    MARITAL_STATUS==hybrid.label~hybrid.label,
    .default = "undecided"
  )) %>% 
  mutate(across(contains('vote'),function(x) replace_na(x,"not given"))) %>% 
  mutate(structured.label=MARITAL_STATUS) %>% 
  mutate(across(contains("label"),function(x) case_when(x=='single person'~'single',
                                                        x=='life partner'~'partner',
                                                        x=='na'~'not given',
                                                        is.na(x)~'not given',
                                                        x==''~'not given',
                                                        x=='unknown (default)'~'not given',
                                                        .default=x))) %>%
  mutate(across((contains('label')&!contains('alone'))&!contains('ref.livingwith')|contains('vote'),~factor(.x)))#,levels=c('single','not given','married','partner','engaged to be married','widowed','separated','divorced','dating','relationship breakdown'))))  


saveRDS(df.14,'df.14')

# df.nursing=df.iii.allnotes.3 %>% 
#   filter(CATEGORY=='Nursing/other') %>% 
#   filter(label=='married') #' only reliable in married
```

We will get the latest discharge summary if there are multiple entries during one admission

```{r}
df.14=readRDS('df.14')
library(tidytable)  #' tidytable speeds up operations
  df.15=
    df.14 %>% 
  #  bind_rows(df.nursing) %>% 
  #  mutate(label.present=ifelse(is.na(label)|label=='not given',0,1)) %>% 
    filter(!is.na(label),label!='not given') %>% 
    group_by(HADM_ID) %>% 
    arrange(desc(CHARTDATE)) %>% 
    slice(1) %>% 
    ungroup %>% 
    select(-contains("ROW_ID"))  
  
detach('package:tidytable') #' tidy fill doesn't work correctly with tidytable

```

The next calculations combine social history notes with metrics such as height, weight, patient details, admission details, A1C, etc. 

```{r}
  require(tidyverse)
  df.16=
  df.15 %>% 
  left_join(elix.iii %>% left_join(elix.iii.2),
            by=c("HADM_ID"="hadm_id")) %>% 
  left_join(echo.iii %>% 
              filter(!is.na(hadm_id)) %>% 
              group_by(hadm_id) %>% 
              arrange(chartdate) %>% 
              slice(1) %>% 
              ungroup 
            ,by=c("HADM_ID"="hadm_id")) %>%  
  #left_join(adm) %>% select(-ROW_ID) %>% 
  left_join(pat) %>% 
  select(-contains('ROW_ID')) %>% 
  left_join(icu.iii) %>% 
  left_join(hw.iii, by=c("ICUSTAY_ID"="icustay_id")) %>% 
  left_join(a1c %>% 
              filter(!is.na(HADM_ID),ITEMID==50852) %>% select(HADM_ID,ITEMID,VALUENUM) %>% 
              distinct(HADM_ID,ITEMID,.keep_all=T), by='HADM_ID')
```

add other metrics: protein and cholesterol levels
if there are multiple measurements in lab events per admission take only one metric per admission

```{r}
  df.17=
  df.16 %>% 
  tidytable::left_join(c.protein %>% select(HADM_ID,VALUENUM) %>% 
                         rename(reactive.protein=VALUENUM) %>% distinct(HADM_ID,.keep_all=T),
                       by='HADM_ID') %>% 
  tidytable::left_join(chol %>% select(HADM_ID,VALUENUM) %>% 
                         rename(cholesterol.total=VALUENUM) %>% distinct(HADM_ID,.keep_all=T),
                       by='HADM_ID') 
```

convert datetime to date format and set weight to first measurement of weight

```{r}
  df.18=
  df.17 %>%   
  mutate(ADMITTIME=lubridate::as_date(ADMITTIME)) %>%   
  mutate(weight=weight_first)
```

delete entries, where there is no marital status detected, calculate patient's age and time between admissions 
delete outliers (aged > 120 and aged < 18)

```{r}
  df.19=
  df.18 %>% 
  filter(label!='not given',label!='na',!is.na(label)) %>% 
  group_by(SUBJECT_ID) %>% 
  arrange(ADMITTIME,.by_group=T) %>% 
  mutate(time=ADMITTIME-tidytable::lag(ADMITTIME)) %>% 
  ungroup %>% 
  mutate(DOB=lubridate::as_date(DOB)) %>% 
  mutate(age=trunc((ADMITTIME-DOB)/365)) %>% 
  mutate(age=as.integer(age)) %>%  
  filter(age<120,age>=18)  
```

we select only married, divorced and single people since UC model showed high precision for these groups only, other groups are not reliable
we select only latest admission per patient

```{r}
  df.20=
  df.19 %>% 
  # filter(label %in% c("married",
  #                     "single person",
  #                     "divorced"
  # )) %>%
  group_by(SUBJECT_ID) %>%
  mutate(admissions=n()) %>%      #' count number of admission per patient
  mutate(more.than.one.admission=ifelse(admissions>1,T,F)) %>%
  arrange(desc(ADMITTIME)) %>%
  slice(1) %>%
  ungroup %>%
  #rename(ADMITDATE=ADMITTIME) %>% 
  #left_join(adm,by='HADM_ID',copy=F) %>%   
  mutate(DISCHTIME=lubridate::as_datetime(DISCHTIME)) %>% 
  mutate(ADMITTIME=lubridate::as_datetime(ADMITTIME)) %>% 
  mutate(days.to.discharge=as.integer(DISCHTIME-ADMITTIME)/(60*24)) %>% 
  # mutate(label=factor(label,c( "single person","partner", "married", 
  #                              "divorced", "separated", "widowed"),ordered=T)) %>%
  mutate(label=fct_drop(label)) 

  df.20$label %>% table
```

bin admissions into groups

```{r}
  df.20.married=df.20 %>%filter(label=='married') 
  df.20.single=df.20 %>%filter(label=='single') 
  df.20.divorced=df.20 %>% filter(label=='divorced')
  df.20.dating=df.20 %>% filter(label=='dating')
  df.20.partner=df.20 %>% filter(label=='partner')
  df.20.widowed=df.20 %>% filter(label=='widowed')
```

we can see that mean ages are quite different , we need to downsample larger groups

```{r}
df.20.married$age %>% mean
df.20.single$age %>% mean
df.20.divorced$age %>% mean
df.20.dating$age %>% mean
df.20.widowed$age %>% mean
df.20.partner$age %>% mean
df.20.dating$age %>% mean

# df.20.married.2=df.20.married %>%
#   mutate(sample_weight=1/age^3.9) %>%
#   slice_sample(n=ceiling(nrow(.)*0.12),weight_by = sample_weight)
# df.20.divorced.2=df.20.divorced %>%
#   mutate(sample_weight=1/age^6) %>%
#   slice_sample(n=ceiling(nrow(.)*0.3),weight_by = sample_weight)
# df.20.widowed.2=df.20.widowed %>%
#   mutate(sample_weight=1/age^40) %>%
#   slice_sample(n=ceiling(nrow(.)*0.1),weight_by = sample_weight)
# df.20.single.2=df.20.single %>%
#   mutate(sample_weight=1/age^5) %>%
#   slice_sample(n=ceiling(nrow(.)*0.8),weight_by = sample_weight)


df.20.married.2=df.20.married %>%
  mutate(sample_weight=1/age^2.5) %>%
  slice_sample(n=ceiling(nrow(.)*0.3),weight_by = sample_weight)
df.20.married.2$age %>% mean
df.20.divorced.2=df.20.divorced %>%
  mutate(sample_weight=1/age^2) %>%
  slice_sample(n=ceiling(nrow(.)*0.8),weight_by = sample_weight)
df.20.divorced.2$age %>% mean
df.20.widowed.2=df.20.widowed %>%
  mutate(sample_weight=1/age^13) %>%
  slice_sample(n=ceiling(nrow(.)*0.08),weight_by = sample_weight)
df.20.widowed.2$age %>% mean
df.20.single.2=df.20.single %>%
  mutate(sample_weight=1/age^-3) %>%
  slice_sample(n=ceiling(nrow(.)*0.8),weight_by = sample_weight)
df.20.single.2$age %>% mean
df.20.dating.2=df.20.dating %>%
  mutate(sample_weight=1/age^-3.5) %>%
  slice_sample(n=ceiling(nrow(.)*0.6),weight_by = sample_weight)
df.20.dating.2$age %>% mean
df.20.partner.2=df.20.partner %>%
  mutate(sample_weight=1/age^-4) %>%
  slice_sample(n=ceiling(nrow(.)*0.6),weight_by = sample_weight)
df.20.partner.2$age %>% mean

car::leveneTest(age~label,bind_rows(df.20.divorced.2,df.20.married.2))
car::leveneTest(age~label,bind_rows(df.20.partner,df.20.married.2))
car::leveneTest(age~label,bind_rows(df.20.widowed.2,df.20.married.2))
car::leveneTest(age~label,bind_rows(df.20.partner.2,df.20.dating.2))
car::leveneTest(age~label,bind_rows(df.20.partner.2,df.20.single.2))


layout(matrix(c(1, 2, 3, 4,5,6), nrow = 2, 
              ncol = 3, byrow = TRUE)) 
df.20.married.2$age %>% hist
df.20.divorced.2$age %>% hist
df.20.widowed.2$age %>% hist
df.20.dating.2$age %>% hist
df.20.partner.2$age %>% hist
df.20.single.2$age %>% hist

df.21=bind_rows(df.20.married.2
                ,df.20.divorced.2
                ,df.20.widowed.2
                ,df.20.dating
                ,df.20.partner
                ,df.20.single.2)
```

Now average age is ~53

```{r}
require(ggpubr)
```

Build boxplot comparing continous health metrics between single , married and divorced groups

```{r}
bxp=
#   bind_rows(
#   df.20.married.2,
#   df.20.divorced.2,
#   df.20.single
# ) %>%
  df.21 %>% 
  mutate(DOB=lubridate::as_date(DOB)) %>%  
  mutate(age=trunc(as.integer(lubridate::as_date(ADMITTIME)-DOB)/365)) %>%   
  mutate(age=as.integer(age)) %>% 
  group_by(SUBJECT_ID) %>% 
  mutate(admissions=n()) %>% 
  arrange(desc(ADMITTIME)) %>% slice(1) %>% ungroup %>% 
  left_join(adm,by='HADM_ID',copy=F)   %>% 
  mutate(DISCHTIME=lubridate::as_datetime(DISCHTIME.x)) %>% 
  mutate(ADMITTIME=lubridate::as_datetime(ADMITTIME.x)) %>% 
  mutate(days.to.discharge=as.integer((DISCHTIME-ADMITTIME)/24)) %>% 
  mutate(weight=ifelse(weight>200|weight<20,NA,weight)) %>% 
  mutate(bpsys=ifelse(bpsys>300,NA,bpsys)) %>% 
  mutate(bpdias=ifelse(bpdias>300,NA,bpdias)) %>% 
  mutate(bpdias=ifelse(bpdias>300,NA,bpdias)) %>% 
  rename(a1c=VALUENUM) %>% 
  pivot_longer(c(#age,
                 admissions,
                 #a1c,
                 #reactive.protein,
                 #weight,
                 #bpsys,bpdias,
                 #cholesterol.total,
                 elixhauser_vanwalraven
                 #days.to.discharge
                 ),
               names_to='vars.orig',values_to='values.orig'
  ) %>%
  ggboxplot(x="label",y="values.orig",fill = "#00AFBB",
            facet.by = "vars.orig",scales = "free")+
  geom_pwc(method='t_test',vjust=2,ref.group=2)

bxp 
ggadjust_pvalue(
  bxp, p.adjust.method = "holm",
  symnum.args = list(cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, Inf), symbols = c("****", "***", "**", "*", "")),
  label = "p adjusted={p.adj.format} {p.adj.signif}", hide.ns = F
)
```

Although there was difference, especially in A1C levels and elixhauser
after adjusting p values no significant changes were detected

```{r}


# bind_rows(
#   df.20.married.2,
#   df.20.divorced.2,
#   df.20.single
# ) %>% 
  df.21 %>% 
  left_join(adm,by='HADM_ID',copy=F)   %>% 
  mutate(DISCHTIME=lubridate::as_datetime(DISCHTIME.x)) %>% 
  mutate(ADMITTIME=lubridate::as_datetime(ADMITTIME.x)) %>% 
  mutate(days.to.discharge=as.integer(DISCHTIME-ADMITTIME)/(60*24)) %>% 
  pivot_longer(c(depression,drug_abuse,alcohol_abuse),
               names_to='vars.orig',values_to='values.orig'
  ) %>%
  drop_na(values.orig) %>% 
  mutate(values.orig=as.logical(values.orig)) %>% 
  group_by(vars.orig,label) %>% add_tally(name='nn') %>% 
  group_by(vars.orig,label,values.orig) %>%  add_tally %>%
  filter(values.orig==T) %>% 
  select(n,nn) %>% distinct %>% 
  mutate(c.min=binom.test(x=n,n=nn)$conf.int[1],c.max=binom.test(x=n,n=nn)$conf.int[2]) %>% 
  mutate(vars.orig=gsub('[_.]',' ',vars.orig)) %>%  
  mutate(label=fct_recode(label,'single'='single person')) %>% 
  filter(vars.orig!='more than one admission') %>% 
    mutate(label=factor(label,levels = c("single", "dating",
                                         "married", "partner", "widowed", "separated", "divorced"))) %>% 
    
  ggplot(aes(y=n*100/nn,x=label))+
  geom_col(width=0.62,aes(fill=label)) + 
  geom_errorbar(aes(ymin=c.min*100,ymax=c.max*100),position='dodge',width = 0.2)+
  ylab('Percentage of patients with the given diagnosis')+
  xlab("")+
  theme_minimal()+
  scale_fill_manual(values=    c("#003f5c",
                                   #"#2f4b7c",
                                   "#665191",
                                   "#a05195",
                                   #"#d45087",
                                   "#f95d6a",
                                   "#ff7c43",
                                   "#ffa600"
    ))+
   theme(legend.position = "none")+
   theme( axis.text.x = element_text( angle = 45 ,
                                      vjust = 1 ,
                                      hjust = 1 ) )+
   theme(aspect.ratio = 2.3/1)+
   facet_grid(~vars.orig,scales='free')
  
  #' We group data frame by patient and calculate difference of marital status and health metrics between admissions
df.21=
  df.19 %>% 
  filter(label %in% c("married",
                       "single person",
                       "divorced",
                       "dating",
                      "widowed",
                      "single",
                      "partner"
  )) %>%
  ungroup %>% 
  group_by(SUBJECT_ID) %>% 
  arrange(ADMITTIME) %>% 
  mutate(time=ADMITTIME-dplyr::lag(ADMITTIME)) %>% 
  mutate(label.prev=dplyr::lag(label)) %>% 
  #distinct() %>% 
  mutate(
    a1c.delta=(VALUENUM-lag(VALUENUM))/lag(VALUENUM),
    reactive.protein.delta=(reactive.protein-lag(reactive.protein))/lag(reactive.protein),
    weight.delta=abs(weight-lag(weight))/lag(weight),
    BP.dias.delta=(bpsys-lag(bpsys))/lag(bpsys),
    BP.sist.delta=(bpdias-lag(bpdias))/lag(bpdias),
    cholesterol.total.delta=(cholesterol.total-lag(cholesterol.total))/lag(cholesterol.total),
    elixhauser_vanwalraven.delta=(elixhauser_vanwalraven-lag(elixhauser_vanwalraven))/lag(elixhauser_vanwalraven)
  ) %>%
  ungroup 
```

calculate more metrics and 'change' column indicating if change in marital status happened

```{r}
df.22=df.21 %>%  
  mutate(DOB=lubridate::as_date(DOB)) %>% 
  mutate(age=as.numeric(trunc((lubridate::as_date(ADMITTIME)-DOB)/365))) %>% 
  group_by(SUBJECT_ID) %>% 
  arrange(ADMITTIME) %>% 
  mutate(     
    depression.delta=depression-lag(depression),  
    alcohol_abuse.delta=alcohol_abuse-lag(alcohol_abuse),
    drug_abuse.delta=drug_abuse-lag(drug_abuse),
    hypertension.delta=hypertension-lag(hypertension),
    metastatic_cancer.delta=metastatic_cancer-lag(metastatic_cancer),
    depression.delta=ifelse(depression.delta==0&depression==1,2,depression.delta),  ### coding schema -1 recovery, 0 - no change, No Dx, 1 - new Dx, 2 - no change, Dx present
    alcohol_abuse.delta=ifelse(alcohol_abuse.delta==0&alcohol_abuse==1,2,alcohol_abuse.delta),
    drug_abuse.delta=ifelse(drug_abuse.delta==0&drug_abuse==1,2,drug_abuse.delta),
    metastatic_cancer.delta=ifelse(metastatic_cancer.delta==0&metastatic_cancer==1,2,metastatic_cancer.delta),
    hypertension.delta=ifelse(hypertension.delta==0&hypertension==1,2,hypertension.delta)
  ) %>% 
  mutate(change=ifelse(label!=label.prev&(label=='married'),"Got married",
                       ifelse(label!=label.prev&label!='married',"Other change",
                              "Unchanged"))) %>% 
  filter(!is.na(change)) %>% 
  ungroup %>% 
  filter(time<=5*365)
  
df.22$change %>% table
# give.n <- function(x){
#   return(c(y = median(x)*1.25, 
#            label = length(x))) 
#   # experiment with the multiplier to find the perfect position
# }


df.22.married=
  df.22 %>%
  filter(change=='Got married') 

df.22.other=
  df.22 %>%
  filter(change=='Other change')  

df.22.unchanged=
  df.22 %>%
  filter(change=='Unchanged')

df.22.change=bind_rows(df.22.married,df.22.other)
```

groups have different age, we need to downsample again

```{r}
df.22.married$age %>% mean
df.22.other$age %>% mean
df.22.unchanged$age %>% mean
df.22.change$age %>% mean


df.22.unchanged.2=
  df.22.unchanged %>% ungroup %>% 
  mutate(sample_weight=age^(-3)) %>% 
  slice_sample(n=2500,weight_by = sample_weight)
t.test(df.22.unchanged.2$age,df.22.change$age)
df.22.unchanged.2$age %>% mean
df.22.unchanged.2$age %>% hist
df.22.change$age %>% hist
var.test(df.22.unchanged.2$age,df.22.change$age)
# df.9.all.younger.51=bind_rows(df.9.unchanged.2,df.9.change) %>% 
#   mutate(sample_weight=age^(-100)) %>% 
#   slice_sample(n=1700,weight_by = sample_weight)
# df.9.all.younger.51$age %>% mean


bind_rows(df.22.unchanged.2,df.22.change) %>% 
  mutate(change=ifelse(change=='Unchanged',change,"Changed")) %>% 
  ungroup %>% 
  filter(time<=5*365) %>% 
  #select(time,contains('label'),everything()) %>% View
  mutate(across(contains("delta")&!contains('depression'),function(x) x*100)) %>%
  pivot_longer(contains('delta'),names_to='vars',values_to = 'values') %>% 
  filter(!(vars %in% c("depression.delta","alcohol_abuse.delta","drug_abuse.delta"))) %>% 
  ggboxplot(x="change",y="values",fill = "#00AFBB",
            facet.by = "vars",scales = "free")+
  geom_pwc(method='t_test',vjust=2)
bxp.change 
ggadjust_pvalue(
  bxp.change, p.adjust.method = "holm",
  
  symnum.args = list(cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, Inf), symbols = c("****", "***", "**", "*", "")),
  label = "p adjusted={p.adj.format} {p.adj.signif}", hide.ns = F
)



  df.22 %>% 
  ungroup %>% 
  filter(!is.na(label.prev),label.prev!='na',!is.na(label),label!='na') %>% 
  distinct(SUBJECT_ID,label,label.prev,.keep_all = T) %>%  
  mutate(change=ifelse(label!=label.prev&label=='married',"Changed",#"Got married",
                       ifelse(label!=label.prev&label!='married',"Changed",#"Other change",
                              "Unchanged"))) %>% pull(change) %>% table
```

Display Changed vs Unchanged chart

```{r}
  bind_rows(df.22.unchanged.2,df.22.change) %>% 
  #filter(time<=5*365) %>%
    mutate(change=ifelse(label!=label.prev&label=='married',"Status changed",
                         ifelse(label!=label.prev&label!='married',"Status changed",
                                "Unchanged"))) %>%
  pivot_longer(contains('delta'),names_to='vars',values_to = 'values') %>% 
  #filter(vars %in% names(df.iii.allnotes.combined.4)[which(grepl("delta",names(df.iii.allnotes.combined.4)))]) %>% 
  group_by(vars,change) %>% add_tally(name='tot') %>% 
  group_by(vars,values,change) %>% add_tally(name='n') %>% 
  select(n,tot) %>% distinct %>% 
  mutate(c.min=binom.test(x=n,n=tot)$conf.int[1],c.max=binom.test(n,tot)$conf.int[2]) %>% 
  mutate(vars=gsub(".delta","",vars)) %>% 
  filter(!is.na(values)) %>% 
  mutate(y=n*100/tot) %>% 
  mutate(values=factor(ordered=T,values,levels=c(-1,0,2,1),labels=c("Recovery/ Dx Removed","No Change in Dx - no Dx",'No change in Dx - Dx present','New Dx'))) %>% 
  mutate(vars=gsub("_"," ",vars)) %>% 
  filter(vars %in% c('alcohol abuse','depression','drug abuse')) %>% 
  #filter(vars %in% c('alcohol abuse','depression','drug abuse','hypertension','metastatic cancer')) %>%   # adhoc tests
  mutate(change=fct_recode(change,"Same Marital Status"="Unchanged","Change in Marital Status"="Status changed")) %>% 
  ggplot(aes(y=n*100/tot,x=values,fill=factor(change))) +
  scale_color_brewer(palette = 'Dark2')+
  geom_col(position='dodge2',color='darkgrey')+
  scale_color_hue(direction = -1)+
  scale_fill_manual(values=c('#88CCEE',"white")) +
  geom_errorbar(color='red',aes(ymin=c.min*100,ymax=c.max*100),position=position_dodge(width=0.9),width=0.2)+
  theme_minimal()+
  theme( axis.text.x = element_text( angle = 45 ,
                                     vjust = 1 ,
                                     hjust = 1 ) )+
  theme(aspect.ratio = 2.1/1)+
  theme(legend.title = element_blank(),legend.position = "top")+
  theme(axis.title = element_text(size = 9))+              # Axis titles
  facet_wrap(~vars)+
  xlab("Comparative Diagnosis (Dx) Status from the First Admission to Second Admission")+
  ylab('Percentage of Patient admission event pairs')
```

Seperate analysis for "Got married" vs "No change" vs "Other change"

```{r}
  bind_rows(df.22.unchanged.2,df.22.change) %>% 
    mutate(change=ifelse(label!=label.prev&(label=='married'),"Got married",
                         ifelse(label!=label.prev&label!='married',"Other change",
                                "Unchanged"))) %>% 
   pivot_longer(contains('delta'),names_to='vars',values_to = 'values') %>%
    #filter(vars %in% c("depression.delta","alcohol_abuse.delta","drug_abuse.delta")) %>% 
    filter(vars %in% c('alcohol_abuse.delta','depression.delta','drug_abuse.delta','hypertension.delta','metastatic_cancer.delta')) %>%   # adhoc tests
    group_by(vars,change) %>% add_tally(name='tot') %>% 
    group_by(vars,values,change) %>% add_tally(name='n') %>% 
    select(n,tot) %>% distinct %>% 
    mutate(c.min=binom.test(x=n,n=tot)$conf.int[1],c.max=binom.test(n,tot)$conf.int[2]) %>% 
    mutate(vars=gsub(".delta","",vars)) %>% 
    filter(!is.na(values)) %>% 
    mutate(y=n*100/tot) %>% 
    mutate(values=factor(ordered=T,values,levels=c(-1,0,2,1),labels=c("Recovery/ Dx Removed","No Change in Dx - no Dx",'No change in Dx - Dx present','New Dx'))) %>% 
    ggplot(aes(y=n*100/tot,x=values,fill=factor(change))) +
    geom_col(position='dodge')+
    geom_errorbar(aes(ymin=c.min*100,ymax=c.max*100),position='dodge')+
    scale_color_hue(direction = -1)+
    scale_fill_manual(values=c('#88CCEE',"white",'black')) +
    facet_wrap(~vars)+theme(legend.title=element_blank())+
    xlab("")+
    ylab('%')

  
  #bxp=
    bind_rows(df.22.unchanged.2,df.22.change) %>% 
    mutate(change=ifelse(change=="Unchanged",change,"Status Change")) %>% 
    # filter(change=="Status Change") %>% select(contains('delta')) %>% 
    mutate(DOB=lubridate::as_date(DOB)) %>%  
    mutate(age=trunc(as.integer(lubridate::as_date(ADMITTIME)-DOB)/365)) %>%   
    mutate(age=as.integer(age)) %>% 
    group_by(SUBJECT_ID) %>% 
    mutate(admissions=n()) %>% 
    arrange(desc(ADMITTIME)) %>% slice(1) %>% ungroup %>% 
    left_join(adm,by='HADM_ID',copy=F)   %>% 
    mutate(DISCHTIME=lubridate::as_datetime(DISCHTIME.x)) %>% 
    mutate(ADMITTIME=lubridate::as_datetime(ADMITTIME.x)) %>% 
    mutate(days.to.discharge=as.integer((DISCHTIME-ADMITTIME)/24)) %>% 
    mutate(weight=ifelse(weight>200|weight<20,NA,weight)) %>% 
    mutate(bpsys=ifelse(bpsys>300,NA,bpsys)) %>% 
    mutate(bpdias=ifelse(bpdias>300,NA,bpdias)) %>% 
    mutate(bpdias=ifelse(bpdias>300,NA,bpdias)) %>% 
    rename(a1c=VALUENUM) %>% 
    pivot_longer(c(#age,
      a1c.delta,
      reactive.protein.delta,
      weight.delta,
      BP.sist.delta,
      BP.dias.delta,
      cholesterol.total.delta,
      elixhauser_vanwalraven.delta
      #days.to.discharge
    ),
    names_to='vars.orig',values_to='values.orig'
    ) %>%
    ggboxplot(x="change",y="values.orig",fill = "#00AFBB",
              facet.by = "vars.orig",scales = "free")+
    geom_pwc(method='t_test',vjust=2,ref.group=2)
  
  bxp 
  #' after adjusting p values no significant changes were detected
  ggadjust_pvalue(
    bxp, p.adjust.method = "holm",
    symnum.args = list(cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, Inf), symbols = c("****", "***", "**", "*", "")),
    label = "p adjusted={p.adj.format} {p.adj.signif}", hide.ns = F
  )
```



---
title: podium_git.R
author: scherbak
date: '2024-01-19'

---
